---
title: "Post-secondary pathways"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-after-ct.csv") %>%
  mutate(ps.grad = ifelse(ps.grad == "Y", "Yes", ps.grad),
         highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Associate degree or lesser cred",
                                     ifelse(highest.cred.level %in% c("Some college", "No college"), "Did not grad ps", highest.cred.level)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "1", "4-year public",
                                            ifelse(ps.grad.InstitutionSector == "4", "2-year public",
                                                   ifelse(ps.grad.InstitutionSector == "11", "Other type of college", ps.grad.InstitutionSector))),
         ps.grad.location = ifelse(ps.grad.location %in% c("PS grad EDR", "PS grad region"), "PS grad region", ps.grad.location))

kable(names(master.original))

kable(head(master.original))
```

# Dependent Variable

The cross-tab and ANOVA analysis exploring relationships between demographic, high school, and post-secondary variables with employment outcomes highlighted the importance of post-secondary variables in an individuals journey to meaningful employment inside and outside the region. Due to this, it makes sense to explore if some of the other variables play a role in post-secondary outcomes - whether an individual attends, graduates, the type of institution they attend and the location.

We will explore this here. We will explore this in three sections;

## Tier 1: Core Postsecondary Completion

These outcomes are the most foundational and are most closely tied to long-term employment and economic stability.

1.  **ps.grad** (graduated from postsecondary)

**Question:** Who completes a postsecondary credential?\
**Why it matters:** Credential completion is consistently the strongest predictor of sustained employment and earnings. This outcome captures whether students translate postsecondary participation into tangible labor market value.

2.  **highest.cred.level**

**Question:** What level of credential do graduates earn?\
**Why it matters:** Certificates, associate degrees, and bachelor’s degrees lead to very different employment opportunities and wage trajectories. This outcome distinguishes not just completion, but the depth of educational attainment.

## Tier 2: Secondary / Stratifying Postsecondary Outcomes

These outcomes add important context by differentiating *how* and *where* postsecondary education occurs.

3.  **ps.grad.InstitutionSector**

**Question:** From what type of institution do students graduate?\
**Why it matters:** Institutional sector (e.g., public two-year, public four-year, private) shapes cost, access, program offerings, and alignment with regional labor markets.

4.  **ps.grad.location** (used when geography matters)

**Question:** Where do students complete their postsecondary education?\
**Why it matters:** Location reveals patterns of mobility, regional retention, and place-based opportunity. It is especially relevant for understanding workforce pipelines and regional equity.

## Tier 3: Early Access Outcome (Optional)

This outcome captures initial engagement with postsecondary education and helps contextualize later attainment.

5.  **attended.ps** (ever attended postsecondary)

**Question:** Who participates in postsecondary education at any point?\
**Why it matters:** This provides a broad view of access beyond immediate enrollment, helping distinguish barriers to entry from barriers to completion.

Each dependent variable has the following categories;

```{r dependent variable categories}
data <- master.original %>%
  select(ps.grad, highest.cred.level, ps.grad.InstitutionSector, ps.grad.location, attended.ps) %>%
  mutate(across(c("ps.grad":"attended.ps"), factor))
```

**`ps.grad`**

```{r echo = FALSE}
levels(data$ps.grad)

```

**`highest.cred.level`**

```{r echo = FALSE}
levels(data$highest.cred.level)
```

`ps.grad.InstitutionSector`

```{r echo = FALSE}
levels(data$ps.grad.InstitutionSector)
```

`ps.grad.location`

```{r echo = FALSE}
levels(data$ps.grad.location)
```

`attended.ps`

```{r echo = FALSE}
levels(data$attended.ps)
```

<br>

# Predictors

For all of these, we will use the predictors that were in Tier 1 and Tier 2 of the previous cross-tab analysis.

-   Demographics
    -   `Gender`
    -   `economic.status`
    -   `RaceEthnicity`
    -   `SpecialEdStatus`
-   High school characteristics
    -   `Dem_Desc`
    -   `edr`
    -   `avg.unemp.rate`
    -   `avg.wage.pct.state`
-   High school accomplishments
    -   `ACTCompositeScore`
    -   `pseo.participant`
    -   `cte.achievement`
    -   `total.cte.courses.taken`
    -   `MCA.M`
    -   `MCA.R`
    -   `MCA.S`

<br>

# [**Tier 1: Core Postsecondary Completion**]{.underline}

## Tier 1 Integrated Summary: Postsecondary Completion and Credential Level

Tier 1 examined two core outcomes: whether students complete postsecondary education (ps.grad) and the highest credential earned (highest.cred.level). Across both outcomes, clear and consistent patterns emerge.

[**Strongest Predictors Across Tier 1**]{.underline}

**PSEO Participation**\
PSEO participation is the single strongest predictor across both completion and credential level. Students who participate in PSEO are dramatically more likely to complete postsecondary education and far more likely to earn a bachelor’s degree. Early exposure to college coursework sharply differentiates both persistence and upward attainment.

**Academic Readiness (ACT and MCA Scores)**\
Academic performance consistently predicts both whether students graduate and how far they go. Higher ACT scores and stronger MCA proficiency in Math, Reading, and Science are strongly associated with completion. These differences become even more pronounced at the bachelor’s level, where academic readiness clearly separates bachelor’s earners from associate earners and non-completers.

**Economic Status**\
Socioeconomic disadvantage is strongly associated with lower completion and lower credential levels. Students eligible for free or reduced-price lunch are far less likely to complete postsecondary education and substantially less likely to earn bachelor’s degrees. Economic gaps widen at higher levels of attainment.

**Special Education Status**\
Students who received special education services are significantly less likely to complete postsecondary education and are very unlikely to earn bachelor’s degrees. The disparity intensifies when examining credential level.

[**Persistent but Moderately Sized Predictors**]{.underline}

**Race/Ethnicity**\
Racial and ethnic disparities are substantial across both outcomes. American Indian, Black, and Hispanic students are far more likely to not graduate, while White and Asian/PI students are more likely to earn bachelor’s degrees. Inequities are particularly visible at higher levels of attainment.

**Gender**\
Women are more likely than men to complete postsecondary education and to earn bachelor’s degrees. The gender gap is most pronounced at the bachelor’s level rather than at lower credential tiers.

**CTE Measures**\
CTE participation and course-taking patterns are associated with different credential pathways. Higher CTE course volume aligns more closely with associate-level attainment, while bachelor’s attainment is associated with fewer CTE courses. These relationships are meaningful but not as strong as academic readiness or PSEO participation.

[**Weak Contextual Predictors**]{.underline}

**Geographic and Economic Context (EDR, Rural-ness, County Unemployment, County Wages)**\
High school contextual characteristics are consistently weak predictors across both completion and credential level. While regional differences are statistically detectable, geographic and local economic conditions explain only a small portion of variation in attainment.

### Overall Tier 1 Pattern

Across both completion and credential level, the dominant drivers of postsecondary success in Northwest Minnesota are **academic preparation and early college exposure**, followed closely by **structural demographic inequities**. Geographic and local economic context play a comparatively minor role.

Importantly, disparities intensify at higher levels of attainment. Predictors that influence completion become even more pronounced when distinguishing bachelor’s degree earners from all other groups. Tier 1 therefore suggests that improving academic readiness and expanding meaningful early college opportunities are central levers for increasing both postsecondary completion and higher-level credential attainment in the region.

<br>

## `ps.grad`

### Demographics

**1. Race/Ethnicity (Cramér’s V = 0.298)**\
Race and ethnicity show the strongest relationship with postsecondary graduation. Completion rates vary widely across groups, with White students having the highest likelihood of graduating and substantially lower completion among American Indian, Black, and Hispanic students. American Indian students are particularly overrepresented among non-completers, indicating pronounced racial and ethnic inequities in postsecondary attainment.

**2. Economic Status (φ = 0.259)**\
Economic disadvantage is strongly associated with lower postsecondary graduation rates. Students eligible for free or reduced-price lunch are far less likely to complete a credential and are disproportionately represented among non-completers, highlighting persistent socioeconomic barriers to postsecondary success.

**3. Special Education Status (φ = 0.254)**\
Special education status is also strongly related to postsecondary graduation. Students who required special education services complete postsecondary credentials at much lower rates than their peers, suggesting substantial structural and support-related challenges in translating secondary education into postsecondary attainment.

**4. Gender (φ = 0.122)**\
Gender has the weakest relationship among the demographic variables examined, though it remains meaningful. Women are more likely than men to complete a postsecondary credential, while men are disproportionately represented among non-completers. Compared to other demographic factors, however, the gender gap is more modest.

::: panel-tabset
#### `Gender`

**Statistical relationship?:** Yes

**Strength of relationship:** Small-to-moderate (φ = 0.122, p \< .001)

**Description of relationship:** Women are more likely than men to complete a postsecondary credential. Just over half of women (50.3%) graduated from postsecondary education, compared with 38.1% of men. Men are disproportionately represented among non-completers, indicating a meaningful gender gap in postsecondary completion in Northwest Minnesota.

```{r}

data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `economic.status`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (φ = 0.259, p \< .001)

**Description of relationship:** Economic status is strongly associated with postsecondary graduation. Individuals who were not eligible for free or reduced-price lunch are much more likely to complete a postsecondary credential (53.9%) than those who were eligible (27.3%). Conversely, nearly three-quarters of economically disadvantaged students do not complete postsecondary education, highlighting substantial and persistent socioeconomic barriers to college completion in Northwest Minnesota.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `RaceEthnicity`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.298, p \< .001)

**Description of relationship:** Postsecondary graduation rates vary substantially by race and ethnicity. White students have the highest completion rate (51.4%), while completion rates are notably lower for American Indian (16.2%), Black (24.4%), Hispanic (25.3%), and Asian/PI students (43.0%). American Indian students in particular are overwhelmingly represented among non-completers. These differences indicate pronounced racial and ethnic inequities in postsecondary completion in Northwest Minnesota, extending well beyond what would be expected by chance alone.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `SpecialEdStatus`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (φ = 0.254, p \< .001)

**Description of relationship:** Special education status is strongly associated with postsecondary graduation. Individuals who did not receive special education services complete postsecondary credentials at much higher rates (49.4%) than those who did (14.8%). Students with special education needs are overwhelmingly represented among non-completers, indicating substantial structural and support-related barriers to postsecondary completion for this population in Northwest Minnesota.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

### High School Characteristics

**1. Economic Development Region (Cramér’s V = 0.086)**\
Economic Development Region shows the strongest relationship among the high school characteristics, though the effect remains small. Students from EDR 4 – West Central have the highest completion rate (47.2%), followed by EDR 1 – Northwest (43.0%), while EDR 2 – Headwaters has the lowest rate (36.0%). Regional context appears to matter, but only modestly.

**2. Rural-ness of School District (Dem_Desc) (Cramér’s V = 0.030)**\
District rural classification is statistically related to postsecondary completion, but the effect is very small. Completion rates vary only slightly across district types, ranging from about 41% to 47%. Rural-ness alone does little to meaningfully explain differences in postsecondary graduation.

**3. Average County Unemployment Rate (ANOVA, p \< .001)**\
Average unemployment during high school years differs statistically between completers and non-completers, but the mean difference is very small (≈ 0.13 percentage points). Local unemployment conditions during high school have minimal practical association with postsecondary completion.

**4. Average County Wages as Percent of State (ANOVA, p \< .001)**\
Average county wage levels during high school years are statistically associated with completion, but the difference is extremely small (≈ 0.01). County-level wage context provides very limited explanatory value for postsecondary graduation outcomes.

**Overall Pattern:**\
Compared to demographic factors, high school contextual characteristics are far weaker predictors of postsecondary completion. While region shows a modest relationship, broader geographic and economic context variables explain only a small portion of variation in who completes a credential in Northwest Minnesota.

<br>

::: panel-tabset
#### `Dem_Desc`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (Cramér’s V = 0.030, p \< .001)

**Description of relationship:** Rural-ness of the school district is statistically associated with postsecondary completion, but the relationship is extremely small. Completion rates range from 41.4% in urban/town/rural mix districts to 46.5% in entirely urban districts, with entirely rural and town/rural mix districts falling in between (about 44%). These differences are minor in magnitude, indicating that district rural classification alone does little to explain variation in postsecondary completion in Northwest Minnesota.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `EDR`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.086, p \< .001)

**Description of relationship:** Economic Development Region is associated with postsecondary completion, though the magnitude of the relationship is small. Students from EDR 4 – West Central have the highest completion rate (47.2%), followed by EDR 1 – Northwest (43.0%), while EDR 2 – Headwaters has the lowest rate (36.0%). Although regional differences are statistically meaningful, region alone explains only a modest share of variation in postsecondary graduation.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `avg.unemp.rate`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county unemployment rate during high school years differs statistically between those who completed postsecondary education and those who did not. Individuals who graduated experienced slightly higher average unemployment rates than non-completers (mean difference ≈ 0.13 percentage points). However, the size of this difference is minimal, indicating that local unemployment conditions during high school have very limited practical influence on postsecondary completion.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

```

#### `avg.wages.pct.state`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county wages (as a percent of the Minnesota average) during high school years differ statistically between completers and non-completers. Students who completed postsecondary education attended high schools in counties with slightly higher relative wages (mean difference ≈ 0.01). However, the magnitude of this difference is extremely small, suggesting that county-level wage context during high school has minimal practical influence on postsecondary completion.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))
```
:::

<br>

### High School Accomplishments

**1. PSEO Participation (φ = 0.299)**\
Participation in postsecondary educational opportunities during high school shows the strongest relationship with postsecondary completion. Students who participated in PSEO graduate at dramatically higher rates (63.1%) than non-participants (31.7%). Early exposure to college-level coursework appears to be one of the most powerful predictors of eventual credential completion.

**2. ACT Composite Score (ANOVA, p \< .001; mean difference ≈ 2.78 points)**\
ACT performance is strongly associated with postsecondary graduation. Students who completed a credential scored nearly three points higher on average than non-completers, indicating that academic preparation and college readiness play a major role in completion outcomes.

**3. MCA Reading (ANOVA, p \< .001; mean difference ≈ 0.51)**\
Reading proficiency shows a substantial relationship with completion. Students who graduated from postsecondary programs scored meaningfully higher in reading than non-completers, underscoring the importance of literacy skills for long-term academic persistence.

**4. MCA Math (ANOVA, p \< .001; mean difference ≈ 0.49)**\
Math proficiency is also strongly associated with completion, with completers scoring nearly half a point higher on average. Academic skill gaps in mathematics appear closely tied to postsecondary success.

**5. MCA Science (ANOVA, p \< .001; mean difference ≈ 0.44)**\
Science proficiency follows a similar pattern, with completers scoring moderately higher than non-completers. While slightly smaller than math and reading gaps, the difference remains meaningful.

**6. CTE Achievement Status (Cramér’s V = 0.085)**\
CTE achievement category shows a small relationship with postsecondary graduation. Students with no CTE participation have the highest completion rates, while concentrators/completers have somewhat lower rates. The differences are statistically meaningful but modest in magnitude.

**7. Total CTE Courses Taken (ANOVA, p \< .001; mean difference ≈ -0.70)**\
The total number of CTE courses taken has a small relationship with completion. Postsecondary completers took slightly fewer CTE courses on average than non-completers. Although statistically significant, the practical difference is limited.

**Overall Pattern:**\
High school accomplishments—particularly indicators of academic readiness and early college exposure—are substantially stronger predictors of postsecondary completion than high school contextual characteristics. PSEO participation and academic performance measures (ACT and MCA scores) stand out as the most influential factors, while CTE-related measures show weaker, though still statistically meaningful, associations.

::: panel-tabset
#### `ACTCompositeScore`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** ACT Composite scores differ substantially between students who completed postsecondary education and those who did not. Students who graduated from postsecondary programs scored, on average, about 2.78 points higher on the ACT than non-completers. This is a meaningful gap, indicating that academic preparation—as measured by ACT performance—is strongly associated with postsecondary completion in Northwest Minnesota.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `pseo.participant`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (φ = 0.299, p \< .001)

**Description of relationship:** Participation in postsecondary educational opportunities (PSEO) during high school is strongly associated with postsecondary completion. Students who participated in PSEO complete a credential at much higher rates (63.1%) than those who did not (31.7%). Conversely, non-participants are disproportionately represented among non-completers. Early exposure to college-level coursework appears to be a powerful predictor of eventual postsecondary graduation in Northwest Minnesota.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `cte.achievement`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.085, p \< .001)

**Description of relationship:** Career and Technical Education (CTE) achievement status is associated with postsecondary completion, though the relationship is modest. Students with no CTE participation have the highest completion rate (50.8%), followed by CTE participants (45.4%), while CTE concentrators or completers have the lowest rate (40.3%). While differences across groups are statistically meaningful, the overall magnitude suggests that CTE status alone is not a strong predictor of postsecondary graduation.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `total.cte.courses.taken`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (ANOVA, p \< .001)

**Description of relationship:** The total number of CTE courses taken differs statistically between postsecondary completers and non-completers. Students who completed a postsecondary credential took, on average, about 0.70 fewer CTE courses than those who did not complete. While this difference is statistically significant, the magnitude is modest, suggesting that overall CTE course volume is only weakly associated with postsecondary graduation.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.M`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** MCA Math scores differ meaningfully between postsecondary completers and non-completers. Students who completed a credential scored, on average, about 0.49 points higher than those who did not complete. This gap is substantial, indicating that math proficiency in high school is strongly associated with postsecondary graduation in Northwest Minnesota.

```{r}
summary <- aov(MCA.M ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

```

#### `MCA.R`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** MCA Reading scores differ substantially between students who completed postsecondary education and those who did not. Completers scored, on average, about 0.51 points higher than non-completers. This represents a meaningful gap, indicating that reading proficiency in high school is strongly associated with postsecondary graduation in Northwest Minnesota.

```{r}
summary <- aov(MCA.R ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

```

#### `MCA.S`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** MCA Science scores differ meaningfully between postsecondary completers and non-completers. Students who completed a credential scored, on average, about 0.44 points higher than those who did not. This indicates that science proficiency in high school is also strongly associated with postsecondary graduation, though the gap is slightly smaller than what was observed in math and reading.

```{r}
summary <- aov(MCA.S ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

### Summary of `ps.grad`

[**Strong and Consistent Predictors**]{.underline}

**PSEO Participation (φ = 0.299)**\
Early exposure to college coursework is one of the strongest predictors of completion. Students who participated in PSEO graduate at dramatically higher rates than non-participants, indicating that structured early access to postsecondary experiences substantially increases the likelihood of earning a credential.

**Academic Performance (ACT and MCA Scores)**\
ACT Composite scores and MCA proficiency in Math, Reading, and Science all show moderate relationships with completion. Students who complete postsecondary education consistently demonstrate higher academic proficiency in high school. Academic readiness is a central driver of postsecondary success.

**Race/Ethnicity (Cramér’s V = 0.298)**\
Racial and ethnic disparities in completion are pronounced. White students complete at substantially higher rates than several other groups, particularly American Indian students. These inequities are large and persistent.

**Economic Status (φ = 0.259)**\
Socioeconomic disadvantage is strongly associated with lower completion. Students eligible for free or reduced-price lunch graduate at far lower rates, underscoring the structural impact of economic barriers.

[**Moderate Predictors**]{.underline}

**Special Education Status (φ = 0.254)**\
Students who required special education services complete at significantly lower rates, suggesting challenges in transitioning from secondary to postsecondary systems.

**CTE Measures (Cramér’s V = 0.085; ANOVA p \< .001)**\
CTE achievement status and total CTE courses taken show small but meaningful relationships with completion. However, their predictive strength is notably weaker than academic readiness and PSEO participation.

[**Weak Contextual Predictors**]{.underline}

**Economic Development Region (Cramér’s V = 0.086)**\
Regional differences exist, but the effect size is small. Region alone explains only a modest share of variation in completion.

**Rural-ness of District (Cramér’s V = 0.030)**\
District rural classification has a statistically detectable but practically minimal relationship with completion.

**County Unemployment and Wage Context (ANOVA p \< .001)**\
Although statistically significant, differences in local economic conditions during high school are extremely small in magnitude and provide little explanatory value for postsecondary graduation.

[**Overall Pattern**]{.underline}

Postsecondary completion in Northwest Minnesota is driven far more by **individual academic readiness and early college exposure** than by geographic or local economic context. Demographic inequities—particularly by race/ethnicity and economic status—remain substantial. High school accomplishments, especially PSEO participation and academic proficiency, emerge as the most influential levers associated with whether students ultimately earn a credential.

<br>

## `highest.cred.level`

### Demographics

**1. Economic Status (Cramér’s V = 0.272)**\
Economic status shows the strongest relationship with highest credential earned. Students not eligible for free or reduced-price lunch are far more likely to earn a bachelor’s degree or higher and far less likely to leave postsecondary without graduating. Economically disadvantaged students are disproportionately concentrated among non-completers, indicating that socioeconomic status strongly shapes both completion and credential level.

**2. Special Education Status (Cramér’s V = 0.266)**\
Special education status is nearly as strong a predictor as economic status. Students who received special education services are overwhelmingly more likely to not graduate from postsecondary education and very unlikely to earn a bachelor’s degree. The disparities extend beyond completion into the level of attainment.

**3. Race/Ethnicity (Cramér’s V = 0.216)**\
Race and ethnicity show a substantial relationship with credential level. White and Asian/PI students are more likely to earn bachelor’s degrees, while American Indian, Black, and Hispanic students are far more likely to not graduate. These differences reflect persistent inequities in both access to and attainment of higher-level credentials.

**4. Gender (Cramér’s V = 0.141)**\
Gender has the weakest relationship among the demographic variables, though it remains meaningful. Women are more likely than men to earn a bachelor’s degree and less likely to leave postsecondary without graduating. The gender gap is most pronounced at the bachelor’s level rather than at the associate degree level.

**Overall Pattern:**\
Demographic factors are strongly associated not only with whether students complete postsecondary education, but also with the level of credential earned. Socioeconomic status and special education status emerge as the most influential demographic predictors of credential level, followed by race/ethnicity. Gender differences are present but comparatively smaller. Across all variables, disparities widen at the bachelor’s degree level, indicating that equity gaps are particularly pronounced at higher levels of attainment.

::: panel-tabset
#### `Gender`

**Statistical relationship?:** Yes

**Strength of relationship:** Small-to-moderate (Cramér’s V = 0.141, p \< .001)

**Description of relationship:** Gender is meaningfully associated with highest credential earned. Women are more likely than men to earn a bachelor’s degree or higher (33.6% vs. 21.5%) and less likely to leave postsecondary without graduating (49.9% vs. 62.0%). Rates of earning an associate degree or lesser credential are nearly identical across genders (16.5%). The primary gender gap appears at the bachelor’s degree level and in overall completion.

```{r}
data <- master.original %>%
  filter(highest.cred.level != "Attending ps")

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `economic.status`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.272, p \< .001)

**Description of relationship:** Economic status is strongly associated with highest credential earned. Students not eligible for free or reduced-price lunch are much more likely to earn a bachelor’s degree or higher (35.6%) compared to economically disadvantaged students (13.4%). Economically disadvantaged students are substantially more likely to not graduate from postsecondary education at all (72.9% vs. 46.3%). Socioeconomic status therefore plays a major role not only in whether students complete, but in the level of credential attained.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `RaceEthnicity`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (Cramér’s V = 0.216, p \< .001)

**Description of relationship:** Highest credential earned varies substantially by race and ethnicity. White and Asian/PI students are far more likely to earn a bachelor’s degree or higher, while American Indian, Black, and Hispanic students are much more likely to not graduate from postsecondary education at all. American Indian students in particular are overwhelmingly represented among non-completers. These patterns indicate pronounced racial and ethnic disparities not only in completion, but in the level of credential attained.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `SpecialEdStatus`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.266, p \< .001)

**Description of relationship:** Special education status is strongly associated with highest credential earned. Students who did not receive special education services are far more likely to earn a bachelor’s degree or higher (31.8%) compared to those who did (3.3%). Students who required special education services are overwhelmingly more likely to not graduate from postsecondary education at all (85.3%). These differences indicate substantial disparities in both completion and level of credential attained for students receiving special education services.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

**1. Economic Development Region (Cramér’s V = 0.066)**\
Economic Development Region shows the strongest relationship among high school contextual variables, though the effect remains small. Students from EDR 4 – West Central have the highest bachelor’s degree attainment and lowest non-completion rates, while EDR 2 – Headwaters has the lowest bachelor’s attainment and highest non-completion rates. Regional context appears to matter modestly for credential level.

**2. Rural-ness of School District (Dem_Desc) (Cramér’s V = 0.040)**\
District rural classification is statistically associated with credential level, but the magnitude is very small. Bachelor’s degree attainment is somewhat higher in entirely urban districts, while non-completion rates are slightly higher in mixed and rural districts. Differences across district types are limited in size.

**3. Average County Unemployment Rate (ANOVA, p \< .001)**\
County unemployment during high school years differs slightly across credential levels, but differences are minimal. Non-completers attended high schools in counties with marginally lower unemployment rates, and bachelor’s and associate degree earners do not meaningfully differ from one another. The practical influence is negligible.

**4. Average County Wages as Percent of State (ANOVA, p \< .001)**\
County wage context is statistically related to credential level, but differences are extremely small. Associate degree earners attended high schools in counties with slightly higher relative wages, while non-completers attended counties with the lowest relative wages. The magnitude of these differences provides little substantive explanatory power.

**Overall Pattern:**\
High school contextual characteristics show weak relationships with highest credential earned. While regional differences are present, geographic and economic context explain only a small portion of variation in credential level. Compared to demographic and academic accomplishment variables, high school characteristics play a limited role in shaping postsecondary attainment levels in Northwest Minnesota.

::: panel-tabset
#### `Dem_Desc`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (Cramér’s V = 0.040, p \< .001)

**Description of relationship:** Rural-ness of the school district is statistically associated with highest credential earned, but the relationship is minimal. Bachelor’s degree attainment is somewhat higher in entirely urban districts (32.4%) compared to entirely rural (24.9%) and urban/town/rural mix districts (25.5%). Differences across district types are modest, and non-completion rates remain similar across categories. District rural classification alone provides very limited explanatory value for variation in credential level.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `EDR`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.066, p \< .001)

**Description of relationship:** Economic Development Region is associated with highest credential earned, though the effect is small. Students from EDR 4 – West Central have the highest rate of bachelor’s degree attainment (29.9%) and the lowest non-completion rate (52.9%), while EDR 2 – Headwaters shows the lowest bachelor’s attainment (23.2%) and the highest non-completion rate (64.3%). Regional differences are statistically meaningful but explain only a modest share of variation in credential level.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `avg.unemp.rate`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county unemployment rate during high school years differs slightly across credential levels. Students who did not graduate from postsecondary education attended high schools in counties with marginally lower unemployment rates than those who earned credentials, while differences between associate degree earners and bachelor’s degree earners are not statistically significant. Although statistically detectable, these differences are minimal in magnitude and provide very limited practical insight into variation in credential level.

```{r}
summary <- aov(avg.unemp.rate ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `avg.wages.pct.state`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county wages (as a percent of the state average) during high school years differ slightly across credential levels. Students who earned associate degrees attended high schools in counties with marginally higher relative wages than both bachelor’s degree earners and non-completers, while non-completers attended high schools in counties with the lowest relative wages. Although statistically significant, the magnitude of these differences is extremely small and offers limited practical explanation for variation in highest credential earned.

```{r}
summary <- aov(avg.wages.pct.state ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

**1. PSEO Participation (Cramér’s V = 0.333)**\
PSEO participation shows the strongest relationship with credential level. Students who participated in PSEO are dramatically more likely to earn a bachelor’s degree or higher and far less likely to leave postsecondary without graduating. Early college exposure sharply differentiates not just completion, but upward movement into higher-level credentials.

**2. ACT Composite Score (ANOVA, p \< .001; large mean differences)**\
ACT scores strongly distinguish bachelor’s degree earners from both associate degree earners and non-completers. Bachelor’s earners scored nearly four points higher than non-completers. Academic readiness, as measured by ACT performance, is a powerful predictor of attaining a bachelor’s degree.

**3. MCA Proficiency (Math, Reading, Science) (ANOVA, p \< .001)**\
MCA Math, Reading, and Science scores all show strong relationships with credential level. Bachelor’s degree earners consistently outperform associate degree earners, who in turn outperform non-completers. Academic proficiency clearly differentiates levels of attainment, especially at the bachelor’s level.

**4. Total CTE Courses Taken (ANOVA, p \< .001)**\
The number of CTE courses taken shows a moderate relationship with credential level. Students earning associate degrees take the most CTE courses on average, while bachelor’s degree earners take substantially fewer. CTE course volume appears aligned with associate-level pathways rather than bachelor’s attainment.

**5. CTE Achievement Status (Cramér’s V = 0.115)**\
CTE achievement category shows a small-to-moderate relationship with credential level. Students without CTE participation are most likely to earn bachelor’s degrees, while concentrators/completers are more likely to earn associate degrees or not complete. CTE pathways appear associated with different postsecondary trajectories.

------------------------------------------------------------------------

**Overall Pattern:**\
High school accomplishments are strongly associated with not only whether students complete postsecondary education, but the level of credential they earn. Academic readiness indicators (PSEO participation, ACT scores, MCA proficiency) are the most powerful differentiators of bachelor’s degree attainment. CTE measures show meaningful but more pathway-specific relationships, particularly aligning with associate-level credentials. Overall, academic preparation and early exposure to college-level coursework emerge as the clearest drivers of higher-level postsecondary attainment in Northwest Minnesota.

::: panel-tabset
#### `ACTCompositeScore`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** ACT Composite scores differ substantially across credential levels. Students who earned a bachelor’s degree or higher scored nearly 3.8 points higher on average than those earning an associate degree or lesser credential, and nearly 4 points higher than those who did not graduate. Differences between associate degree earners and non-completers are minimal and not statistically significant. Academic readiness, as measured by ACT performance, is strongly associated with earning a bachelor’s degree and clearly differentiates higher levels of attainment.

```{r}
summary <- aov(ACTCompositeScore ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `pseo.participant`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (Cramér’s V = 0.333, p \< .001)

**Description of relationship:** PSEO participation is strongly associated with highest credential earned. Students who participated in PSEO are far more likely to earn a bachelor’s degree or higher (45.3%) compared to non-participants (16.2%), and far less likely to leave postsecondary without graduating (37.0% vs. 68.4%). Early exposure to college coursework sharply differentiates not only completion, but progression to higher-level credentials.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `cte.achievement`

**Statistical relationship?:** Yes

**Strength of relationship:** Small-to-moderate (Cramér’s V = 0.115, p \< .001)

**Description of relationship:** CTE achievement status is meaningfully associated with highest credential earned. Students with no CTE participation are the most likely to earn a bachelor’s degree or higher (38.7%), while CTE concentrators/completers are more likely to earn an associate degree or lesser credential (19.1%) and less likely to earn a bachelor’s degree (20.9%). Non-completion rates are highest among CTE concentrators/completers. These patterns suggest CTE pathways are associated with different postsecondary trajectories, with bachelor’s attainment more common among students not concentrated in CTE.

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `total.cte.courses.taken`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** The total number of CTE courses taken differs meaningfully across credential levels. Students who earned a bachelor’s degree or higher took substantially fewer CTE courses than those earning an associate degree or lesser credential (difference ≈ 1.79 courses). Non-completers also took fewer CTE courses than associate degree earners but more than bachelor’s degree earners. These patterns suggest that higher volumes of CTE coursework are more closely aligned with associate-level pathways, while bachelor’s attainment is associated with lower CTE course participation.

```{r}
summary <- aov(total.cte.courses.taken ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.M`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Math scores differ substantially across credential levels. Students who earned a bachelor’s degree or higher scored about 0.51 points higher than associate degree earners and nearly 0.69 points higher than non-completers. Associate degree earners also scored meaningfully higher than non-completers. Math proficiency clearly differentiates levels of postsecondary attainment, with higher performance strongly associated with bachelor’s degree completion.

```{r}
summary <- aov(MCA.M ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.R`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Reading scores vary substantially by highest credential earned. Students who earned a bachelor’s degree or higher scored about 0.46 points higher than associate degree earners and roughly 0.68 points higher than non-completers. Associate degree earners also scored meaningfully higher than non-completers. Reading proficiency strongly differentiates levels of postsecondary attainment, particularly distinguishing bachelor’s degree earners from all other groups.

```{r}
summary <- aov(MCA.R ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.S`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Science scores differ substantially across credential levels. Students who earned a bachelor’s degree or higher scored about 0.46 points higher than associate degree earners and roughly 0.61 points higher than non-completers. Associate degree earners also scored meaningfully higher than non-completers. Science proficiency strongly differentiates higher levels of postsecondary attainment, particularly distinguishing bachelor’s degree earners from other groups.

```{r}
summary <- aov(MCA.S ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

### Summary of `highest.cred.level`

[**Strong and Consistent Predictors**]{.underline}

**PSEO Participation (Cramér’s V = 0.333)**\
Early college participation is the strongest predictor of credential level. Students who participated in PSEO are far more likely to earn a bachelor’s degree and far less likely to leave postsecondary without graduating. Early exposure to college coursework strongly differentiates higher-level attainment.

**Academic Readiness (ACT and MCA Scores)**\
ACT Composite scores and MCA proficiency in Math, Reading, and Science all show strong relationships with credential level. Bachelor’s degree earners consistently outperform associate degree earners, who in turn outperform non-completers. Academic preparation clearly separates levels of postsecondary attainment, particularly at the bachelor’s level.

**Economic Status (Cramér’s V = 0.272)**\
Socioeconomic status strongly shapes credential level. Students not eligible for free or reduced-price lunch are substantially more likely to earn bachelor’s degrees and far less likely to not graduate. Economic disadvantage is closely tied to lower levels of attainment.

**Special Education Status (Cramér’s V = 0.266)**\
Students who required special education services are overwhelmingly more likely to not graduate and very unlikely to earn bachelor’s degrees. Special education status remains a major differentiator of credential level.

[**Moderate Predictors**]{.underline}

**Race/Ethnicity (Cramér’s V = 0.216)**\
Racial and ethnic disparities persist across credential levels. White and Asian/PI students are more likely to earn bachelor’s degrees, while American Indian, Black, and Hispanic students are more likely to not graduate. Gaps widen at higher levels of attainment.

**Gender (Cramér’s V = 0.141)**\
Women are more likely than men to earn bachelor’s degrees and less likely to leave postsecondary without graduating. The gender gap is most pronounced at the bachelor’s level.

**Total CTE Courses Taken (ANOVA, p \< .001)**\
CTE course volume differentiates credential pathways. Associate degree earners take the most CTE courses, while bachelor’s degree earners take substantially fewer, suggesting pathway specialization.

**CTE Achievement Status (Cramér’s V = 0.115)**\
CTE category is associated with different postsecondary trajectories, with concentrators/completers more likely to earn associate-level credentials and less likely to earn bachelor’s degrees.

[**Weak Contextual Predictors**]{.underline}

**Economic Development Region (Cramér’s V = 0.066)**\
Regional differences in credential level exist but are modest. Region explains only a small share of variation in attainment.

**Rural-ness of District (Cramér’s V = 0.040)**\
District rural classification has a very small relationship with credential level.

**County Unemployment and Wage Context (ANOVA, p \< .001)**\
Local economic conditions during high school are statistically related to credential level but the differences are extremely small and offer little practical explanatory value.

[**Overall Pattern**]{.underline}

Highest credential earned in Northwest Minnesota is driven primarily by academic readiness and early college exposure, with substantial demographic inequities shaping attainment levels. Academic performance and PSEO participation sharply differentiate bachelor’s degree earners from other groups. Geographic and local economic context, while statistically significant, play a comparatively minor role in explaining variation in credential level.

<br>

# [**Tier 2: Post-secondary stratification**]{.underline}

## Tier 2 Integrated Summary: Institutional Sector and Graduation Location

Tier 2 examined two stratifying outcomes: the type of institution from which students graduate (ps.grad.InstitutionSector) and where they graduate (ps.grad.location). Together, these outcomes show how postsecondary pathways differentiate not just completion, but *tier of institution* and *geographic mobility*.

[**Strongest Predictors Across Tier 2**]{.underline}

**PSEO Participation**\
PSEO participation is the strongest and most consistent predictor across both institutional sector and graduation location. Students who participate in PSEO are dramatically more likely to graduate from 4-year public institutions and far less likely to not graduate. They are also substantially more likely to graduate outside the region. Early college exposure functions as a structural accelerator into bachelor’s-oriented and more mobile pathways.

**Academic Readiness (ACT and MCA Scores)**\
Academic performance strongly differentiates both sector and location. Students with higher ACT scores and stronger MCA proficiency are far more likely to graduate from 4-year public institutions and more likely to graduate outside the region. Non-completers consistently show the lowest academic performance. Academic readiness is the primary sorting mechanism shaping both institutional tier and geographic mobility.

**Economic Status**\
Socioeconomic status strongly shapes Tier 2 outcomes. Students not eligible for free or reduced-price lunch are far more likely to graduate from 4-year institutions and far more likely to graduate beyond the region. Economically disadvantaged students are disproportionately concentrated among non-completers and less mobile graduates. Socioeconomic disparities influence both attainment level and geographic reach.

**Special Education Status**\
Students who received special education services are overwhelmingly more likely to not graduate and are far less likely to access 4-year or out-of-region pathways. Disparities are substantial across both institutional tier and graduation location.

[**Persistent but Moderately Sized Predictors**]{.underline}

**Race/Ethnicity**\
Racial and ethnic disparities remain evident across sector and location. White and Asian/PI students are more likely to graduate from 4-year institutions and to graduate outside the region, while American Indian, Black, and Hispanic students are more concentrated among non-completers. Disparities extend beyond completion into institutional stratification and mobility.

**Gender**\
Women are more likely than men to graduate from 4-year institutions and more likely to graduate within or beyond Minnesota. Men are more concentrated among non-completers. Gender differences are meaningful but smaller than socioeconomic or special education disparities.

**CTE Measures**\
CTE achievement and course volume align more strongly with 2-year public institutional pathways and somewhat with regional retention. Students with higher CTE intensity are more likely to graduate from 2-year institutions and less likely to graduate from 4-year institutions or outside the region. These patterns are meaningful but not as strong as academic readiness or PSEO participation.

[**Weak Contextual Predictors**]{.underline}

**Geographic and Economic Context (EDR, Rural-ness, County Unemployment, County Wages)**\
High school contextual characteristics are consistently weak predictors across both institutional sector and graduation location. While regional differences are statistically detectable, geographic classification and county-level economic conditions explain only a small share of variation in postsecondary stratification and mobility.

### Overall Tier 2 Pattern

Across both institutional sector and graduation location, the dominant drivers of stratification in Northwest Minnesota are **academic preparation and early college exposure**, followed closely by **structural demographic inequities**.

Students with strong academic readiness and access to PSEO are far more likely to enter 4-year public institutions and to graduate beyond the region. Students facing socioeconomic disadvantage or special education barriers are disproportionately concentrated in non-completion and lower-tier or less mobile pathways. High school context contributes comparatively little explanatory power.

Tier 2 shows that postsecondary opportunity is not only about completion—it is also about institutional tier and geographic mobility, and those pathways are structured by readiness and equity gaps.

<br>

## `ps.grad.InstitutionSector`

### Demographics

**1. Economic Status (Cramér’s V = 0.271)**\
Economic status shows the strongest relationship with institution sector. Students not eligible for free or reduced-price lunch are substantially more likely to graduate from 4-year public institutions and other college types, while economically disadvantaged students are overwhelmingly more likely to not graduate at all. Socioeconomic disparities are especially pronounced in 4-year public pathways.

**2. Special Education Status (Cramér’s V = 0.267)**\
Special education status is nearly as strong a predictor as economic status. Students who received special education services are dramatically less likely to graduate from 4-year public institutions and far more likely to not complete postsecondary education. Disparities are particularly severe at the 4-year level.

**3. Race/Ethnicity (Cramér’s V = 0.176)**\
Race and ethnicity show a moderate relationship with institution sector. White and Asian/PI students are more likely to graduate from 4-year public institutions, while American Indian, Black, and Hispanic students are more likely to not graduate. Sector differences reflect broader completion disparities and uneven access to bachelor’s-level pathways.

**4. Gender (Cramér’s V = 0.149)**\
Gender shows a small-to-moderate relationship with institution sector. Women are more likely than men to graduate from 4-year public institutions and other college types, while men are more likely to not complete postsecondary education. Differences are less pronounced in 2-year public institutions.

**Overall Pattern:**\
Demographic disparities are strongly reflected in postsecondary institution sector outcomes. Socioeconomic status and special education status are the most powerful demographic predictors of sector-level pathways, particularly in access to and completion of 4-year public institutions. Racial/ethnic and gender differences are also meaningful, with disparities most visible in bachelor’s-level institutional sectors. Across variables, inequities in sector pathways largely mirror broader patterns of postsecondary completion and attainment.

::: panel-tabset
#### Gender

**Statistical relationship?:** Yes

**Strength of relationship:** Small-to-moderate (Cramér’s V = 0.149, p \< .001)

**Description of relationship:** Gender is meaningfully associated with postsecondary institution sector. Women are more likely than men to graduate from 4-year public institutions (22.5% vs. 15.3%) and from other types of colleges (13.1% vs. 7.2%), while men are more likely to not graduate from postsecondary education (62.0% vs. 49.9%). Graduation rates from 2-year public institutions are similar across genders. The primary gender differences emerge in bachelor’s-level pathways and overall completion.

```{r}
data <- master.original %>%
  filter(ps.grad.InstitutionSector != "Attending ps") %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector %in% c("Did not grad", "Never attended ps"), "Did not grad ps", ps.grad.InstitutionSector),
         ps.grad.InstitutionSector = fct_relevel(ps.grad.InstitutionSector, "2-year public", "4-year public", "Other type of college", "Did not grad ps"))

tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `economic.status`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.271, p \< .001)

**Description of relationship:** Economic status is strongly associated with postsecondary institution sector. Students not eligible for free or reduced-price lunch are far more likely to graduate from 4-year public institutions (24.5% vs. 9.1%) and other types of colleges (12.6% vs. 5.6%). Economically disadvantaged students are substantially more likely to not graduate from postsecondary education at all (72.9% vs. 46.3%). Socioeconomic disparities are especially pronounced in access to and completion of 4-year public institutions.

```{r}

tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `RaceEthnicity`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (Cramér’s V = 0.176, p \< .001)

**Description of relationship:** Postsecondary institution sector varies meaningfully by race and ethnicity. White and Asian/PI students are more likely to graduate from 4-year public institutions, while American Indian, Black, and Hispanic students are more likely to not graduate from postsecondary education. American Indian students are particularly overrepresented among non-completers. Differences are also visible across 2-year and other institution types, indicating that racial and ethnic disparities extend beyond completion into sector-level pathways.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `SpecialEdStatus`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.267, p \< .001)

**Description of relationship:** Special education status is strongly associated with postsecondary institution sector. Students who did not receive special education services are far more likely to graduate from 4-year public institutions (21.9% vs. 2.1%) and other institution types (11.6% vs. 1.7%). Students who received special education services are overwhelmingly more likely to not graduate from postsecondary education (85.3%). Disparities are particularly pronounced in access to and completion of 4-year institutions.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

**1. Economic Development Region (Cramér’s V = 0.072)**\
Economic Development Region shows the strongest relationship among high school contextual variables, though the effect remains small. EDR 4 – West Central has the highest representation in 4-year public pathways and the lowest non-completion rate, while EDR 2 – Headwaters shows the highest non-completion rate and lower 4-year participation. Regional differences are present but modest in magnitude.

**2. Rural-ness of School District (Dem_Desc) (Cramér’s V = 0.037)**\
District rural classification is statistically associated with institution sector, but the relationship is very small. Entirely urban districts show slightly higher 4-year public graduation rates, while non-completion rates are relatively similar across district types. Rural-ness alone explains little variation in sector outcomes.

**3. Average County Unemployment Rate (ANOVA, p \< .001)**\
County unemployment during high school years differs slightly across institutional sectors, but differences are minimal. Non-completers attended high schools in counties with marginally lower unemployment rates, and there are no meaningful differences among 2-year, 4-year, and other graduates. Practical explanatory value is negligible.

**4. Average County Wages as Percent of State (ANOVA, p \< .001)**\
County wage context is statistically related to institution sector, but the magnitude is extremely small. Students graduating from 2-year public institutions attended counties with slightly higher relative wages, while non-completers attended counties with slightly lower wages. These differences provide very limited insight into sector pathways.

**Overall Pattern:**\
High school contextual characteristics are weak predictors of postsecondary institution sector. While regional differences are modestly detectable, rural classification and county-level economic conditions explain very little variation in institutional pathways. Compared to demographic and academic accomplishment variables, high school characteristics play a limited role in shaping where students ultimately graduate.

::: panel-tabset
#### `Dem_Desc`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (Cramér’s V = 0.037, p \< .001)

**Description of relationship:** Rural-ness of the school district is statistically associated with postsecondary institution sector, but the relationship is minimal. Students from entirely urban districts are slightly more likely to graduate from 4-year public institutions, while non-completion rates are fairly similar across district types. Differences across rural classifications are modest, indicating that district rural-ness provides very limited explanation for variation in institutional sector outcomes.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `edr`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.072, p \< .001)

**Description of relationship:** Economic Development Region is associated with postsecondary institution sector, though the effect is small. EDR 4 – West Central shows the highest graduation rates from 4-year public institutions and the lowest non-completion rates, while EDR 2 – Headwaters has the highest non-completion rate and lower representation in 4-year pathways. Regional differences are statistically meaningful but explain only a modest share of variation in institutional sector outcomes.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE)

```

#### `avg.unemp.rate`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county unemployment rate during high school years differs slightly by postsecondary institution sector. Students who did not graduate from postsecondary education attended high schools in counties with marginally lower unemployment rates than graduates of 2-year, 4-year, or other institution types. Differences among graduates of 2-year, 4-year, and other colleges are not statistically significant. Although statistically detectable, these differences are minimal and provide little practical explanation for variation in institutional sector outcomes.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

```

#### `avg.wages.pct.state`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county wages (as a percent of the state average) during high school years differ slightly by postsecondary institution sector. Students who graduated from 2-year public institutions attended high schools in counties with marginally higher relative wages than students in other sectors, while non-completers attended counties with the lowest relative wages. Although these differences are statistically significant, the magnitude is extremely small and offers limited practical insight into institutional sector outcomes.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

**1. PSEO Participation (Cramér’s V = 0.332)**\
PSEO participation is the strongest predictor of institutional sector. Students who participated in PSEO are dramatically more likely to graduate from 4-year public institutions and other college types, and far less likely to not complete postsecondary education. Early college exposure sharply differentiates 4-year pathways from 2-year and non-completion outcomes.

**2. ACT Composite Score (ANOVA, p \< .001; large mean differences)**\
ACT scores strongly distinguish sector pathways. Students graduating from 4-year public and other institutions score nearly four points higher on average than 2-year public graduates and non-completers. Academic readiness clearly separates 4-year institutional trajectories from 2-year and non-completion outcomes.

**3. MCA Proficiency (Math, Reading, Science) (ANOVA, p \< .001)**\
MCA scores in Math, Reading, and Science all show strong relationships with institution sector. Students graduating from 4-year public institutions consistently outperform 2-year graduates and non-completers. Academic proficiency strongly differentiates bachelor’s-oriented institutional pathways from others.

**4. Total CTE Courses Taken (ANOVA, p \< .001)**\
CTE course volume shows a moderate relationship with sector. Students graduating from 2-year public institutions take the most CTE courses on average, while 4-year public graduates take substantially fewer. CTE intensity aligns more closely with 2-year pathways.

**5. CTE Achievement Status (Cramér’s V = 0.119)**\
CTE achievement category is meaningfully associated with sector. CTE concentrators/completers are more likely to graduate from 2-year public institutions, while students with no CTE participation are more likely to graduate from 4-year institutions. CTE pathways appear aligned with specific postsecondary sectors rather than uniformly predictive of completion.

------------------------------------------------------------------------

**Overall Pattern:**\
High school accomplishments—particularly academic readiness and PSEO participation—are the strongest predictors of postsecondary institution sector. Academic performance sharply differentiates 4-year public pathways from 2-year and non-completion outcomes. CTE measures align more closely with 2-year institutional trajectories. Compared to high school contextual factors, academic preparation and early college exposure play a far more decisive role in shaping where students ultimately graduate.

::: panel-tabset
#### `ACTCompositeScore`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** ACT Composite scores differ substantially by postsecondary institution sector. Students who graduated from 4-year public institutions scored nearly 3.8 points higher on average than 2-year public graduates and nearly 4 points higher than non-completers. Students graduating from other types of colleges show similarly high ACT scores. There is no meaningful difference between 2-year public graduates and non-completers. Academic readiness strongly differentiates 4-year institutional pathways from 2-year and non-completion outcomes.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `pseo.participant`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (Cramér’s V = 0.332, p \< .001)

**Description of relationship:** PSEO participation is strongly associated with postsecondary institution sector. Students who participated in PSEO are far more likely to graduate from 4-year public institutions (31.1% vs. 11.3%) and other college types (15.9% vs. 6.0%), and far less likely to not graduate (37.0% vs. 68.4%). Early college exposure sharply differentiates 4-year institutional pathways from 2-year and non-completion outcomes.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `cte.achievement`

**Statistical relationship?:** Yes

**Strength of relationship:** Small-to-moderate (Cramér’s V = 0.119, p \< .001)

**Description of relationship:** CTE achievement status is meaningfully associated with postsecondary institution sector. CTE concentrators/completers are more likely to graduate from 2-year public institutions, while students with no CTE participation are more likely to graduate from 4-year public institutions and other college types. Non-completion rates are highest among CTE concentrators/completers. These patterns suggest that CTE pathways are more closely aligned with 2-year institutional outcomes, while 4-year pathways are more common among students not concentrated in CTE.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `total.cte.courses.taken`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** The total number of CTE courses taken differs meaningfully by postsecondary institution sector. Students who graduate from 2-year public institutions take the most CTE courses on average, followed by non-completers, while graduates of 4-year public and other institutions take substantially fewer CTE courses. The largest gaps are between 2-year public graduates and 4-year public graduates (difference ≈ 1.74 courses) and between 2-year public and other institution graduates (difference ≈ 2.04 courses). CTE course volume appears strongly aligned with 2-year institutional pathways rather than 4-year outcomes.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.M`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Math scores differ substantially by postsecondary institution sector. Students who graduate from 4-year public institutions score about 0.52 points higher than 2-year public graduates and nearly 0.69 points higher than non-completers. Graduates of other institution types also outperform 2-year graduates and non-completers. Math proficiency clearly differentiates 4-year pathways from 2-year and non-completion outcomes.

```{r}
summary <- aov(MCA.M ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.R`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Reading scores differ substantially by postsecondary institution sector. Students who graduate from 4-year public institutions score about 0.47 points higher than 2-year public graduates and nearly 0.68 points higher than non-completers. Graduates of other institution types show similarly strong performance compared to 2-year graduates. Reading proficiency strongly differentiates 4-year pathways from 2-year and non-completion outcomes.

```{r}
summary <- aov(MCA.R ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

```

#### `MCA.S`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Science scores differ substantially by postsecondary institution sector. Students who graduate from 4-year public institutions score about 0.47 points higher than 2-year public graduates and roughly 0.61 points higher than non-completers. Graduates of other institution types also outperform 2-year graduates and non-completers. Science proficiency clearly differentiates 4-year institutional pathways from 2-year and non-completion outcomes.

```{r}
summary <- aov(MCA.S ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

### Summary

The previous section examines where students graduate—2-year public, 4-year public, other institutions, or not at all. Across groups of predictors, clear patterns emerge.

[**Strong and Consistent Predictors**]{.underline}

**PSEO Participation (Cramér’s V = 0.332)**\
PSEO participation is the strongest predictor of institutional sector. Students who participated in PSEO are dramatically more likely to graduate from 4-year public institutions and far less likely to not complete postsecondary education. Early college exposure sharply differentiates bachelor’s-oriented pathways from 2-year and non-completion outcomes.

**Academic Readiness (ACT and MCA Scores)**\
ACT Composite scores and MCA proficiency in Math, Reading, and Science strongly differentiate sector pathways. Students graduating from 4-year public institutions consistently outperform 2-year public graduates and non-completers. Academic preparation clearly separates 4-year institutional trajectories from other outcomes.

**Economic Status (Cramér’s V = 0.271)**\
Socioeconomic status strongly shapes institutional sector. Students not eligible for free or reduced-price lunch are far more likely to graduate from 4-year public institutions, while economically disadvantaged students are disproportionately represented among non-completers. Disparities are especially visible in access to 4-year public pathways.

**Special Education Status (Cramér’s V = 0.267)**\
Students who received special education services are dramatically less likely to graduate from 4-year institutions and overwhelmingly more likely to not complete postsecondary education. Sector-level disparities mirror and intensify completion gaps.

[**Moderate Predictors**]{.underline}

**Race/Ethnicity (Cramér’s V = 0.176)**\
Racial and ethnic disparities extend into sector pathways. White and Asian/PI students are more likely to graduate from 4-year public institutions, while American Indian, Black, and Hispanic students are more likely to not graduate. Sector differences reflect broader inequities in postsecondary attainment.

**Gender (Cramér’s V = 0.149)**\
Women are more likely than men to graduate from 4-year public institutions and less likely to not complete postsecondary education. Differences are less pronounced in 2-year public pathways.

**CTE Measures (Cramér’s V = 0.119; ANOVA p \< .001)**\
CTE achievement and course volume are associated with sector alignment. Students with higher CTE participation are more likely to graduate from 2-year public institutions, while 4-year public graduates take fewer CTE courses on average. CTE pathways appear aligned with 2-year institutional trajectories.

[**Weak Contextual Predictors**]{.underline}

**Economic Development Region (Cramér’s V = 0.072)**\
Regional differences in sector outcomes are statistically detectable but modest. Region explains only a small share of variation in institutional pathways.

**Rural-ness of District (Cramér’s V = 0.037)**\
District rural classification has a very small relationship with institutional sector.

**County Unemployment and Wage Context (ANOVA, p \< .001)**\
Local economic conditions during high school are statistically related to institutional sector, but differences are extremely small and offer little practical explanatory value.

[**Overall Pattern**]{.underline}

Postsecondary institution sector in Northwest Minnesota is shaped primarily by academic readiness, early college exposure, and structural demographic inequities. Academic performance and PSEO participation sharply differentiate 4-year public pathways from 2-year and non-completion outcomes. Socioeconomic and special education disparities remain substantial, particularly in access to and completion of 4-year institutions. In contrast, high school contextual characteristics play only a minor role in determining institutional sector outcomes.

<br>

## `ps.grad.location`

### Demographics

**1. Economic Status (Cramér’s V = 0.266)**\
Economic status shows the strongest relationship with graduation location. Students not eligible for free or reduced-price lunch are substantially more likely to graduate within the region, elsewhere in Minnesota, and outside Minnesota, while economically disadvantaged students are overwhelmingly more likely to not graduate at all. Socioeconomic disparities are evident in both completion and geographic mobility.

**2. Special Education Status (Cramér’s V = 0.255)**\
Special education status is also a strong predictor of graduation location. Students who received special education services are far more likely to not graduate and far less likely to complete postsecondary education in any geographic location. Disparities extend beyond completion into regional retention and mobility.

**3. Race/Ethnicity (Cramér’s V = 0.173)**\
Race and ethnicity show a moderate relationship with graduation location. White students are more likely to graduate within the region or elsewhere in Minnesota, while American Indian, Black, and Hispanic students are more likely to not graduate. Asian/PI students show relatively higher rates of graduation outside Minnesota. These patterns indicate that racial disparities shape both attainment and mobility outcomes.

**4. Gender (Cramér’s V = 0.125)**\
Gender has the weakest relationship among the demographic variables, though it remains meaningful. Women are more likely than men to graduate within the region and elsewhere in Minnesota and are less likely to not graduate. Differences in out-of-state graduation are smaller. The primary gender gap appears in regional retention and overall completion.

**Overall Pattern:**\
Demographic disparities strongly influence postsecondary graduation location. Socioeconomic status and special education status are the most powerful predictors, shaping not only whether students complete postsecondary education but also where they do so. Racial and gender differences are also meaningful, particularly in regional retention patterns. Across variables, inequities in geographic mobility largely mirror broader disparities in postsecondary completion.

::: panel-tabset
#### Gender

**Statistical relationship?:** Yes

**Strength of relationship:** Small-to-moderate (Cramér’s V = 0.125, p \< .001)

**Description of relationship:** Gender is meaningfully associated with postsecondary graduation location. Women are more likely than men to graduate within the region (25.9% vs. 19.1%) and elsewhere in Minnesota (10.9% vs. 7.5%), and are less likely to not graduate at all (49.9% vs. 62.0%). Differences in graduating outside Minnesota are smaller but still present. The primary gender gap appears in regional retention and overall completion.

```{r}
data <- master.original %>%
  filter(ps.grad.location != "Attending ps") %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "PS grad region", "PS grad MN", "PS grad outside MN", "Did not graduate ps"))

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `economic.status`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.266, p \< .001)

**Description of relationship:** Economic status is strongly associated with postsecondary graduation location. Students not eligible for free or reduced-price lunch are substantially more likely to graduate within the region (26.0%), elsewhere in Minnesota (11.3%), or outside Minnesota (16.4%) compared to economically disadvantaged students (16.3%, 5.4%, and 5.5%, respectively). Economically disadvantaged students are overwhelmingly more likely to not graduate at all (72.9%). Socioeconomic disparities are evident not only in completion but also in geographic mobility of postsecondary attainment.

```{r}

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `RaceEthnicity`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (Cramér’s V = 0.173, p \< .001)

**Description of relationship:** Postsecondary graduation location varies meaningfully by race and ethnicity. White students are far more likely to graduate within the region or elsewhere in Minnesota, while American Indian, Black, and Hispanic students are substantially more likely to not graduate from postsecondary education at all. Asian/PI students show comparatively higher rates of graduation outside Minnesota. These patterns indic

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `SpecialEdStatus`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate-to-strong (Cramér’s V = 0.255, p \< .001)

**Description of relationship:** Special education status is strongly associated with postsecondary graduation location. Students who did not receive special education services are far more likely to graduate within the region (24.8%), elsewhere in Minnesota (10.3%), or outside Minnesota (14.1%), compared to students who received special education services (9.5%, 2.7%, and 2.6%, respectively). Students who received special education services are overwhelmingly more likely to not graduate at all (85.3%). Disparities are evident not only in completion but also in geographic mobility of postsecondary attainment.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

**1. Economic Development Region (Cramér’s V = 0.076)**\
Economic Development Region shows the strongest relationship among high school contextual variables, though the effect remains small. EDR 1 – Northwest and EDR 4 – West Central have similar rates of regional graduation, while EDR 2 – Headwaters shows lower regional graduation and higher non-completion. Graduation outside Minnesota is somewhat more common among students from EDR 4. Regional differences are present but modest.

**2. Rural-ness of School District (Dem_Desc) (Cramér’s V = 0.053)**\
District rural classification is statistically associated with graduation location, but the relationship is small. Entirely urban districts show slightly higher rates of graduation outside Minnesota, while town/rural mix districts show somewhat higher graduation within Minnesota. Differences across district types are limited in magnitude.

**3. Average County Unemployment Rate (ANOVA, p \< .001)**\
County unemployment during high school years differs modestly across graduation locations. Students graduating outside Minnesota and non-completers attended high schools in counties with slightly lower unemployment rates than regional graduates. Differences between regional graduates and those graduating elsewhere within Minnesota are negligible.

**4. Average County Wages as Percent of State (ANOVA, p \< .001)**\
County wage context is statistically related to graduation location, but the differences are extremely small. Students graduating outside Minnesota attended counties with marginally higher relative wages, while non-completers attended counties with slightly lower wages. Practical explanatory value is minimal.

**Overall Pattern:**\
High school contextual characteristics are weak predictors of postsecondary graduation location. While regional differences are statistically detectable, rural classification and county-level economic conditions explain only a small share of variation in geographic graduation outcomes. Compared to demographic and academic accomplishment variables, high school characteristics play a limited role in determining where students ultimately graduate.

::: panel-tabset
#### `Dem_Desc`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.053, p \< .001)

**Description of relationship:** Rural-ness of the school district is statistically associated with postsecondary graduation location, though the relationship is small. Students from entirely urban districts are slightly more likely to graduate outside Minnesota, while town/rural mix districts show somewhat higher rates of graduation elsewhere within Minnesota. Regional graduation rates are fairly similar across district types, and non-completion rates remain consistent. Overall, district rural classification plays only a modest role in explaining where students graduate.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `edr`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.076, p \< .001)

**Description of relationship:** Economic Development Region is associated with postsecondary graduation location, though the relationship is small. EDR 1 – Northwest and EDR 4 – West Central show similar rates of graduating within the region (23.5%), while EDR 2 – Headwaters has a lower regional graduation rate (17.7%) and the highest non-completion rate (64.3%). Graduation outside Minnesota is more common among students from EDR 4. While regional differences are statistically meaningful, they explain only a modest share of variation in geographic graduation outcomes.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE)

```

#### `avg.unemp.rate`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (ANOVA, p \< .001)

**Description of relationship:** Average county unemployment rate during high school years differs modestly by postsecondary graduation location. Students who graduated outside Minnesota attended high schools in counties with lower average unemployment rates than those who graduated within the region, and non-completers show similarly lower unemployment averages compared to regional graduates. Differences between regional graduates and those graduating elsewhere within Minnesota are not statistically significant. Although statistically meaningful, the magnitude of differences is modest.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

```

#### `avg.wages.pct.state`

**Statistical relationship?:** Yes

**Strength of relationship:** Very small (ANOVA, p \< .001)

**Description of relationship:** Average county wages (as a percent of the state average) during high school years differ slightly by postsecondary graduation location. Students who graduated outside Minnesota attended high schools in counties with marginally higher relative wages than regional graduates, while non-completers attended counties with slightly lower relative wages. Differences between regional graduates and those graduating elsewhere within Minnesota are not statistically significant. Although statistically detectable, the magnitude of these differences is extremely small and offers limited practical explanation for geographic graduation outcomes.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

**1. PSEO Participation (Cramér’s V = 0.304)**\
PSEO participation is the strongest predictor of graduation location. Students who participated in PSEO are substantially more likely to graduate within the region, elsewhere in Minnesota, and outside Minnesota, and far less likely to not graduate at all. Early college exposure strongly differentiates both completion and geographic mobility.

**2. ACT Composite Score (ANOVA, p \< .001; large mean differences)**\
ACT scores strongly distinguish graduation locations. Students graduating elsewhere in Minnesota score over 2.4 points higher than regional graduates, and those graduating outside Minnesota also score substantially higher. Non-completers score dramatically lower than all graduate groups. Academic readiness is closely tied to both completion and out-of-region mobility.

**3. MCA Proficiency (Math, Reading, Science) (ANOVA, p \< .001)**\
MCA scores in Math, Reading, and Science show strong relationships with graduation location. Students graduating outside the region consistently outperform regional graduates, and non-completers score substantially lower than all graduate groups. Academic proficiency differentiates not only completion, but patterns of geographic mobility.

**4. Total CTE Courses Taken (ANOVA, p \< .001)**\
CTE course volume shows a moderate relationship with graduation location. Students graduating elsewhere in Minnesota or outside Minnesota take fewer CTE courses than regional graduates, while non-completers take the most. Higher CTE intensity is more closely aligned with regional retention or non-completion, while lower CTE participation is associated with mobility.

**5. CTE Achievement Status (Cramér’s V = 0.077)**\
CTE achievement status has a small relationship with graduation location. Differences across CTE groups are modest, with slight variations in regional retention and mobility. Compared to academic readiness measures, CTE pathways play a much smaller role in explaining geographic graduation outcomes.

------------------------------------------------------------------------

**Overall Pattern:**\
High school accomplishments—especially academic readiness and PSEO participation—are the strongest predictors of postsecondary graduation location. Students with higher academic performance are not only more likely to complete postsecondary education, but also more likely to graduate outside the region. CTE measures show moderate alignment with regional retention patterns, but their influence is notably weaker than academic indicators. Compared to high school contextual characteristics, academic preparation plays a far more decisive role in shaping geographic mobility of postsecondary attainment.

::: panel-tabset
#### `ACTCompositeScore`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** ACT Composite scores differ substantially by postsecondary graduation location. Students who graduated elsewhere in Minnesota scored about 2.46 points higher than regional graduates, while those graduating outside Minnesota scored about 1.87 points higher than regional graduates. Non-completers scored significantly lower than all graduate groups—over 4 points lower than those graduating elsewhere in Minnesota. Academic readiness strongly differentiates not only completion, but geographic mobility of postsecondary attainment.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `pseo.participant`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (Cramér’s V = 0.304, p \< .001)

**Description of relationship:** PSEO participation is strongly associated with postsecondary graduation location. Students who participated in PSEO are substantially more likely to graduate within the region (29.7%), elsewhere in Minnesota (14.0%), and outside Minnesota (19.3%) compared to non-participants. Non-participants are far more likely to not graduate at all (68.4% vs. 37.0%). Early college exposure strongly differentiates both completion and geographic mobility of postsecondary attainment.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `cte.achievement`

**Statistical relationship?:** Yes

**Strength of relationship:** Small (Cramér’s V = 0.077, p \< .001)

**Description of relationship:** CTE achievement status is statistically associated with postsecondary graduation location, though the relationship is small. Students with no CTE participation are somewhat more likely to graduate elsewhere in Minnesota or outside Minnesota, while CTE concentrators/completers are slightly more concentrated in regional graduation pathways. Non-completion rates remain highest among CTE concentrators/completers. Overall, CTE pathway differences in geographic graduation patterns are modest compared to academic readiness or demographic factors.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

#### `total.cte.courses.taken`

**Statistical relationship?:** Yes

**Strength of relationship:** Moderate (ANOVA, p \< .001)

**Description of relationship:** The total number of CTE courses taken differs meaningfully by postsecondary graduation location. Students who graduated elsewhere in Minnesota took about 0.83 fewer CTE courses than regional graduates, and those graduating outside Minnesota took about 0.54 fewer courses. Non-completers took more CTE courses than all graduate groups, with differences exceeding one full course compared to graduates elsewhere in Minnesota. Higher CTE course volume is more closely aligned with regional graduation or non-completion, while lower CTE participation is associated with greater geographic mobility.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.M`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Math scores differ substantially by postsecondary graduation location. Students who graduated outside Minnesota scored about 0.28 points higher than regional graduates, while those graduating elsewhere in Minnesota scored about 0.22 points higher. Non-completers scored significantly lower than all graduate groups—over 0.59 points lower than those graduating elsewhere in Minnesota. Math proficiency strongly differentiates not only completion, but geographic mobility of postsecondary attainment.

```{r}
summary <- aov(MCA.M ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```

#### `MCA.R`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Reading scores differ substantially by postsecondary graduation location. Students who graduated elsewhere in Minnesota or outside Minnesota scored about 0.20 points higher than regional graduates, while non-completers scored significantly lower than all graduate groups—over 0.61 points lower than those graduating elsewhere in Minnesota. Differences between graduates within and outside Minnesota are not statistically significant. Reading proficiency strongly differentiates completion and is associated with greater geographic mobility.

```{r}
summary <- aov(MCA.R ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

```

#### `MCA.S`

**Statistical relationship?:** Yes

**Strength of relationship:** Strong (ANOVA, p \< .001)

**Description of relationship:** MCA Science scores differ substantially by postsecondary graduation location. Students who graduated elsewhere in Minnesota or outside Minnesota scored about 0.25 points higher than regional graduates, while non-completers scored significantly lower than all graduate groups—over 0.56 points lower than those graduating elsewhere in Minnesota. Differences between graduates within and outside Minnesota are not statistically significant. Science proficiency strongly differentiates completion and is associated with greater geographic mobility.

```{r}
summary <- aov(MCA.S ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)
```
:::

<br>

### Summary

Tier 2 (Location) examines where students graduate—within the region, elsewhere in Minnesota, outside Minnesota, or not at all. Across predictor groups, clear themes emerge.

[**Strong and Consistent Predictors**]{.underline}

**PSEO Participation (Cramér’s V = 0.304)**\
PSEO participation is one of the strongest predictors of graduation location. Participants are far more likely to graduate in any location and far less likely to not graduate at all. Early college exposure strongly differentiates both completion and geographic mobility.

**Academic Readiness (ACT and MCA Scores)**\
ACT Composite scores and MCA proficiency in Math, Reading, and Science strongly differentiate graduation locations. Students graduating outside the region or elsewhere in Minnesota consistently score higher than regional graduates, while non-completers score substantially lower than all graduate groups. Academic preparation is tightly linked not only to completion, but to geographic mobility.

**Economic Status (Cramér’s V = 0.266)**\
Socioeconomic status strongly shapes graduation location. Students not eligible for free or reduced-price lunch are far more likely to graduate within or beyond the region and far less likely to not graduate at all. Socioeconomic disparities are evident in both attainment and mobility.

**Special Education Status (Cramér’s V = 0.255)**\
Students who received special education services are overwhelmingly more likely to not graduate and far less likely to complete postsecondary education in any location. Disparities extend beyond completion into reduced geographic mobility.

[**Moderate Predictors**]{.underline}

**Race/Ethnicity (Cramér’s V = 0.173)**\
Race and ethnicity meaningfully shape graduation location. White students are more likely to graduate within the region or elsewhere in Minnesota, while American Indian, Black, and Hispanic students are more likely to not graduate. Asian/PI students show comparatively higher rates of graduation outside Minnesota. Disparities influence both attainment and geographic movement.

**Gender (Cramér’s V = 0.125)**\
Women are more likely than men to graduate within the region and elsewhere in Minnesota and are less likely to not graduate. Differences in out-of-state graduation are smaller. Gender gaps are present but less pronounced than socioeconomic or special education disparities.

**CTE Course Volume (ANOVA, p \< .001)**\
CTE intensity shows a moderate relationship with graduation location. Students graduating outside the region tend to take fewer CTE courses, while non-completers take more. CTE pathways appear somewhat aligned with regional retention patterns.

[**Weak Contextual Predictors**]{.underline}

**Economic Development Region (Cramér’s V = 0.076)**\
Regional differences in graduation location are statistically detectable but modest. EDR 2 – Headwaters shows higher non-completion and lower regional graduation, but overall explanatory power is limited.

**Rural-ness of District (Cramér’s V = 0.053)**\
District rural classification has a small relationship with graduation location.

**County Unemployment and Wage Context (ANOVA, p \< .001)**\
Local economic conditions during high school are statistically associated with graduation location, but differences are extremely small and offer little practical explanation for mobility patterns.

[**Overall Pattern**]{.underline}

Postsecondary graduation location in Northwest Minnesota is shaped primarily by academic readiness, early college exposure, and structural demographic inequities. Students with stronger academic preparation are more likely to complete postsecondary education and more likely to graduate outside the region. Socioeconomic and special education disparities significantly constrain both attainment and mobility. In contrast, high school contextual characteristics contribute little to explaining where students ultimately graduate.

<br>

# Overall Synthesis: Postsecondary Pathways in Northwest Minnesota

## 1. Cross-Tier Synthesis (Completion, Credential Level, Sector, and Location)

Across Tier 1 and Tier 2 outcomes, a clear structure emerges in how postsecondary pathways unfold in Northwest Minnesota. Whether examining completion (ps.grad), credential level (highest.cred.level), institutional sector (ps.grad.InstitutionSector), or graduation location (ps.grad.location), the same drivers consistently shape outcomes.

Students with stronger academic preparation and early exposure to college are far more likely to complete postsecondary education, earn bachelor’s degrees, graduate from 4-year public institutions, and do so outside the region. Conversely, students facing socioeconomic disadvantage, special education barriers, or weaker academic readiness are disproportionately concentrated among non-completers and 2-year institutional pathways, with more limited geographic mobility.

Importantly, stratification intensifies as outcomes become more selective. Gaps widen when moving from completion to bachelor’s attainment and from 2-year to 4-year institutional pathways. Tier 2 outcomes reinforce and amplify patterns observed in Tier 1.

## 2. Ranked Predictors Across All Outcomes

**Strongest and Most Consistent Predictors Across All Outcomes** - PSEO participation\
- ACT Composite score and MCA proficiency

These variables consistently and powerfully differentiate completion, credential level, institutional tier, and geographic mobility.

**Strong and Structural Predictors** - Economic status\
- Special education status\
- Race/Ethnicity

These predictors reflect persistent structural inequities that influence not only whether students complete postsecondary education, but also how far they go and where they graduate.

**Moderate / Pathway-Aligned Predictors** - CTE achievement\
- Total CTE courses taken

CTE participation aligns more strongly with 2-year public pathways and regional retention patterns than with bachelor’s attainment or out-of-region mobility.

**Weak Contextual Predictors** - Economic Development Region (EDR)\
- Rural-ness of school district\
- County unemployment\
- County wage levels

High school contextual and local economic characteristics are statistically detectable but explain comparatively little variation in postsecondary outcomes.

## 3. Core Findings

Three overarching conclusions emerge from the full analysis:

1.  Academic readiness is the dominant sorting mechanism.\
    Students with stronger academic preparation consistently access higher-tier institutions, earn higher credentials, and demonstrate greater geographic mobility.

2.  Early college exposure is the most powerful policy lever.\
    PSEO participation dramatically shifts outcomes upward across every postsecondary measure examined.

3.  Structural inequities shape both attainment and mobility.\
    Socioeconomic disadvantage, special education status, and racial disparities meaningfully constrain both completion and access to higher-tier and more mobile postsecondary pathways.

Postsecondary outcomes in the region are not random—they are patterned, predictable, and strongly structured by readiness and access.

## 4. Implications for Policy and Regional Strategy

The findings suggest several strategic considerations:

-   Expanding access to rigorous academic preparation and strengthening readiness pipelines may have the largest long-term impact on both completion and bachelor’s production.
-   Increasing equitable access to PSEO could serve as a high-leverage intervention for shifting students into higher-tier and more mobile pathways.
-   Addressing socioeconomic and special education disparities must remain central to any strategy aimed at increasing regional attainment.
-   CTE pathways play an important role in supporting 2-year and locally aligned workforce outcomes, but they are not substitutes for academic preparation when the goal is bachelor’s attainment.
-   Geographic context alone does not explain postsecondary outcomes; student-level preparation and structural inequities matter far more than rural classification or local labor market conditions.

Taken together, this analysis indicates that improving academic readiness and expanding equitable access to early college opportunities are the most direct levers for strengthening postsecondary success and long-term workforce outcomes in Northwest Minnesota.

# Core questions

## Most Important Variables Related to Graduating Within the Region

The variables most strongly and consistently associated with graduating within the region were:

**PSEO Participation**\
Early college exposure dramatically increased the likelihood of postsecondary completion, which in turn increased the likelihood of graduating within the region. PSEO is one of the most powerful levers influencing regional graduation outcomes.

**Academic Readiness (ACT and MCA Scores)**\
Students with stronger academic preparation were far more likely to complete postsecondary education, including within the region. While the very highest scorers were more geographically mobile, academic readiness clearly separated regional graduates from non-completers.

**Economic Status**\
Socioeconomic disadvantage substantially reduced the likelihood of regional graduation, largely through its strong relationship with non-completion.

**Special Education Status**\
Students who received special education services were far less likely to graduate in any location, including within the region. This was one of the strongest structural constraints observed.

## Most Important Variables Related to 2-Year / Associate-Level Pathways

The variables most strongly associated with graduating from a 2-year institution and/or earning an associate degree or lesser credential were:

**Academic Readiness (ACT and MCA Scores)**\
Academic preparation strongly sorted students into institutional tiers. Students in 2-year pathways had significantly lower academic performance than those in 4-year public institutions, though higher than non-completers.

**PSEO Participation**\
Students who did not participate in PSEO were much more likely to follow 2-year pathways or not complete. PSEO participation strongly shifted students toward bachelor’s attainment.

**CTE Participation and Course Volume**\
Higher CTE course-taking aligned closely with 2-year public institutional pathways and associate-level attainment. CTE intensity was one of the clearest pathway-alignment variables.

**Economic Status and Special Education Status**\
Structural inequities were also evident in 2-year pathways. Economically disadvantaged students and students receiving special education services were more concentrated in associate-level or non-completion outcomes than in bachelor’s pathways.
